Building the Meta Single Cell Transcriptome Ocular Meta-Atlas

### 2021-03-25 manuscript_00.Rmd
text updates

---
title: 'Building the Million Transcriptome Ocular Meta-Atlas'
author:
  - Vinay Swamy:
      institute: bg
  - Temesgen Fufa:
      institute: mgog
  - Robert Hufnagel:
      institute: mgog
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: references.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "The development of highly scalable single cell transcriptome technology has allowed for a high resolution view into huge numbers of single cell transcriptomes. Analyzing the transcriptomes between different projects is highly desirable as this would theoretically allow for better assessment of which effects are consistent across independent studies. However it is difficult to compare and contrast data across different projects as there are substantial batch effects from computational processing, single cell technology utilized, and the natural biological variation. While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method works best. We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing method / parameter combinations. We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge 750,000 cells from 32 datsets and three species to create a massive ocular single cell transcriptome meta-atlas. This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
keywords: "RNA-seq, retina, RPE, ocular, eye, Snakemake, singlecell, scRNA, single-cell, singlecell, scPOP, R"
---

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(pool)
library(RSQLite)
library(formattable)
library(ggalluvial)
library(scattermore)
library(grid)
library(ComplexHeatmap)
library(viridis)
library(Matrix)
library(MatrixGenerics)
library(matrixStats)
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
#tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

qc <- readr::read_tsv('~/data/scEiaD/QC.tsv.gz')

# cell type prediction
load('~/data/scEiaD/predictions_n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.Rdata')

@@

# kallisto stats
run_info <- read_tsv('~/data/scEiaD/aggregated_run_info.tsv.gz')
write_tsv(run_info %>% 
            filter(sample_accession %in% meta_filter$sample_accession) %>% 
            select(-file) %>% 
            left_join(meta_filter %>% 
                        group_by(sample_accession) %>% 
                        summarise(nCount_RNA = mean(nCount_RNA), nFeature_RNA = mean(nFeature_RNA), percent_mt = mean(percent_mt)) %>% 
                        select(sample_accession, nCount_RNA, nFeature_RNA, percent_mt)), 
          file = 'supplemental_file__kallisto_stats.tsv')
```

@@

## The retina, single cell gene analysis, and a recent explosion

@@

The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields. As of January 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study past 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020]. The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015] and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].

@@

While the gene - cell count tables are often made available in repositories like the Gene Expression Omnibus (GEO), there are no requirements to uniformly process the data. This means the count tables cannot be used in cross-study comparisons as even small differences in the computational pipeline (aligner, transcriptome reference, etc.) create study-specific effects. This issue can be addressed only by re-quantifying the data in a uniform environment. Fortunately, due to the continued development of computationally light-weight gene quantification tools into the single-cell space (e.g kallisto bustools, alevin-fry), re-quantification does not require massive compute resources [@melstedModularEfficientPreprocessing2019; @srivastavaAlevinEfficientlyEstimates2019].

@@

We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas. As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc.). The benchmarking system we developed uses a wide variety of metrics that combine in the R package scPOP (single-cell Pick Optimal Parameters). The scEiaD will be of utility to two communities. First, the ocular community who can both search scEiaD for gene expression across many dimensions (e.g. cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and rich automatic cell labelling. Second, the computational community can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment. As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base, the meta-atlas in several data formats, and propose general guidelines to optimally create custom meta-atlases.

@@

Gene-level counts were quantified with the kallisto bustool pseudo-aligner for both the droplet and well based samples. After empty droplet removal, we had `r format(qc %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells. We then created a Seurat v3 object and removed cells which had more than 10% mitochondrial reads across all gene counts or fewer than 250 unique genes quantified. After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells.

@@

To visualize the interplay between batch mixing and cell type distinction we plot the batch mixing LISI score (which has been multiplied by -1) on plotted on the y-axis (higher is better) against the cluster LISI on the x-axis (higher is better). The best performer on both metrics will be in the top right corner  (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')`a). In the same manner we plot the silhouette metric (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')` b).

<!-- ## scPOP reveals different priorities across the different batch integration methods -->

On one extreme we have combat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a). The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each nearly study is has a distinct space in the 2D (`r fig_cap(name = 'figBenchmark', display = 'cite')`b). In our scEiaD dataset we see that like combat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity. Scanorama, fastMNN, and bbknn more prioritize cleanly separating the clusters. With our scEiaD meta-atlas insct, trVAE, and magic do not perform particularly well in batch mixing or cluster purity.

@@

To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI). In this scaled metric that weighs the different benchmarking algorithms evenly, scVI performs best. As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c). We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods. We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance. These results demonstrate the importance of assessing performance in a rigorous manner across many parameters.

@@

```{r figBenchmark, fig.width=14, fig.height=9, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a. Example of a method (combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type). b. no batch correction cleanly separates cell types but does not mix batches (colored by study). c. sumZScale (higher is better) for each method across a variety of data normalizations. All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions. Each color is a different method. d. Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 50 latent dimensions.')

@@

To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: HVG, latent dimensions, and k nearest neighbors. Furthermore we used a recent advance in scVI capability (>= version 0.8.0) adapted from scArches that allows one to build a reference model and project the new cells onto it [@QueryReferenceSinglecell]. We built a human-only scVI model and then projected the mouse and macaque data onto it. We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters. Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data (`r fig_cap(name = 'figBenchmark', display = 'cite')`d). We found the optimal overall parameters to be 5000 HVG, 8 latent dimensions, and with 5 k-nearest neighbors for the cluster finding. We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.1 and 50 nearest neighbors.

@@

As a brief case study of our ML performance, we look at the cell type labels assigned to a Shekhar et al. study where they used SMART-seq on a retinal bipolar cell enriched population of cells [@shekharComprehensiveClassificationRetinal2016] . Even though our ML algorithm was trained only on droplet-based data, our algorithm labels most of this dataset as retinal bipolar cells, with the next most common cell types being Amacrine,  like Shekhkar et al. labeled them (`r fig_cap(name = 'supFigShekharSMART', display = 'cite')`). To more generally evaluate performance we use the precision recall (PR) curve, which visualizes the ability of the model to precisely label known cells at a given confidence. The area under the PR curve (AUC) summarizes the effectiveness of the model across different cell types, with 1 being the highest performance. The xgboost model can predict rods, bipolar cells, and Müller glia with near perfect performance (`r fig_cap(name = 'supFig_PR', display = 'cite')`). Most of the remaining cell types can be predicted with an AUC over 0.95 `r supTab_cap(name = 'supTab_AUCxgboost', display = 'cite')`. The cell types that were the most challenging for the model to predict were the artery, choriocapillaris, and vein. Artery, choriocapillaris, and vein are constructed from endothelial cells and we find that for all three of these, endothelial was the second most common label. The ML model also performs under 0.9 for predicting the neurogenic cell type label; however the most next common label that is predicts are the highly related retinal progenitor populations (RPCs) (`r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`). Overall, our xgboost based ML model shows strong accuracy across the major cell types of the retina (overall  AUC of 0.98).

<!-- ## Certain retina cell types have more stable community consensus IDs -->

<!-- With the wide diversity of the scEiaD we can get a sense on how consistent the retina community is at labelling the major cell types of the retina. By calculating the AUC against each individual cell type and study, we can see which cell types are consistently labeled correctly across most (`r fig_cap(name = 'supFig_PR_cap', display = 'cite')`). Most studies show very similar performance with our ML model. The study with the lowest correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.   -->

```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE, warning = FALSE}
figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'a. Top genes that are differentially expressed across the major cell types of the retina. Genes are colored by which cell type they are differentially expressed in. The dot size is proportional to the percentage of those cells that have detectable levels of the gene. The color of the dot is the log2 scaled CPM expression. b. 2D UMAP projection of scEiaD, colored by cell type. The Tabula Muris data is colored all in gray. c. Facet plot that demonstrates how each major cell type of the retina is contained within a distinct space. d. Confusion matrix of cell type prediction performance of our xgboost labeller between predicted (x axis) and known (y axis) using data withheld from the machine learner. Most of the cell types are indeed labelled as their true type. e. Faceting of 2D UMAP by species and colored by cell type demonstrate how the major cell types of the retina share space with like cell types, despite being from mouse, human, and macaque.')

@@

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 1
input[['pt_size_meta']] <- 1
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['label_toggle']] <- 2
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- NULL
ctp <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)

@@

After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type. For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labeled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`). Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean() %>% round()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean() %>% round()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean() %>% round()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean() %>% round()` after transferring the cell type labels.

As another check on the quality of the cell type assignments, we ran the cell and cluster independent haystack gene search and pairwise differential expression tests between the predicted cell types (see methods for further details). We show the five most consistently differentially expressed genes for each of the major retina cell types (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a). As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene. We expect most of the genes identified to be known in the literature. The first search is the more precise "gene AND cell type" (e.g. "PDE6H AND Cones") and the second search is the more inclusive "gene AND retina" (e.g. "PDE6H AND Retina"). Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (`r supTab_cap(name = 'supTab_topMarkers', display = 'cite')`) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search. The 50 genes had a mean of `r top_markers %>% filter(Gene %in% dplot_markers) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search). In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 2.5 x 10^-12).

@@

## The scVI-based scEiaD UMAP projection blends batches while separating cell types

>>>>

---
title: 'Building the Meta Single Cell Transcriptome Ocular Meta-Atlas'
author:
  - Vinay Swamy:
      institute: bg
  - Temesgen Fufa:
      institute: mgog
  - Robert Hufnagel:
      institute: mgog
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: references.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "The development of highly scalable single cell transcriptome technology has resulted in the creation of thousands of datasets, over 30 in the retina alone. Analyzing the transcriptomes between different projects is highly desirable as this would allow for better assessment of which biological effects are consistent across independent studies. However it is difficult to compare and contrast data across different projects as there are substantial batch effects from computational processing, single cell technology utilized, and the natural biological variation. While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method functions works best. We developed a lightweight R package (scPOP) that brings in batch integration methods and uses a simple heuristic to balance batch merging and celltype/cluster purity. We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge 766,615 cells from 34 retina datsets and three species to create a massive ocular single cell transcriptome meta-atlas. This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
keywords: "RNA-seq, retina, RPE, ocular, eye, Snakemake, singlecell, scRNA, single-cell, singlecell, scPOP, R"
---

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(pool)
library(RSQLite)
library(formattable)
library(ggalluvial)
library(scattermore)
library(grid)
library(ComplexHeatmap)
library(viridis)
library(Matrix)
library(MatrixGenerics)
library(matrixStats)
library(velociraptor)
library(SingleCellExperiment)
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
#tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

srt <- data.table::fread('~/git/scEiaD/data/sample_run_layout_organism_tech.tsv')
qc <- data.table::fread('~/data/scEiaD/QC.tsv.gz')
qc <- qc %>% left_join(srt %>% select(sample_accession, SX = Source) %>% unique(), by = 'sample_accession') %>% filter(SX %in% c('iPSC','Tissue'))
# cell type prediction
load('~/data/scEiaD/predictions_n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.Rdata')

@@

# kallisto stats
run_info <- read_tsv('~/data/scEiaD/aggregated_run_info.tsv.gz')
write_tsv(run_info %>% 
            filter(sample_accession %in% meta_filter$sample_accession) %>% 
            select(-file) %>% 
            left_join(meta_filter %>% 
                        group_by(sample_accession) %>% 
                        summarise(nCount_RNA = mean(nCount_RNA), nFeature_RNA = mean(nFeature_RNA), percent_mt = mean(percent_mt)) %>% 
                        select(sample_accession, nCount_RNA, nFeature_RNA, percent_mt)), 
          file = 'supplemental_file__kallisto_stats.tsv')

# write top markers to sup file
write_tsv(top_markers, file = 'supplemental_file__celltype_markers.tsv')

# load mini human velocity
load('~/data/scEiaD/velo_scEiaD_2021-03-18_basemodel_homosapiens_tiny.Rdata')
umap_mat <- reducedDim(velo_sceX, 'X_scviUMAP')
grid.df <- gridVectors(umap_mat, reducedDim(velo_sceX, 'velocity_scviUMAP'))

# load ML species transfer data
# created with src/organism_celltype_ML.R
load('~/data/scEiaD/species_transfer_celltype_model.Rdata')
```

@@

## A plethora of single-cell transcriptome studies in the retina

@@

The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields. As of early 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study starting in 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020]. The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015]. As of late 2020, over a twenty published studies, cumulatively containing over a million cells, have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].

@@

While the gene - cell count tables are generally made available in repositories like the Gene Expression Omnibus (GEO), there are no requirements to uniformly process the data. This means the count tables cannot be used in cross-study comparisons as even small differences in the computational pipeline (aligner, transcriptome reference, etc.) create study-specific effects. This issue can be addressed only by re-quantifying the data in a uniform computational environment. Fortunately, due to the continued development of computationally light-weight gene quantification tools in the single-cell space (e.g kallisto bustools, alevin-fry), re-quantification does not require massive compute and time resources [@melstedModularEfficientPreprocessing2019; @srivastavaAlevinEfficientlyEstimates2019].

@@

We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas of retina single cell transcriptomes. As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc.). The benchmarking system we developed uses a wide variety of metrics that combine in the R package scPOP (single-cell Pick Optimal Parameters). The scEiaD will be of utility to two communities. First, the ocular community who can both search scEiaD for gene expression across many dimensions (e.g. cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and rich automatic cell labelling. Second, the computational community can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment. As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base, the meta-atlas in several data formats, and propose general guidelines to optimally create custom meta-atlases.

@@

Gene-level counts were quantified with the kallisto bustools pseudo-aligner for both the droplet and well based samples. After empty droplet removal, we had `r format(qc %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells. We then removed cells which had more than 10% mitochondrial reads across all gene counts or fewer than 200 unique genes quantified. After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells (`r supFig_cap(name = 'supFig1', display = 'cite')`).

@@

To visualize the interplay between batch mixing and cell type distinction we plot the batch mixing LISI score (which has been multiplied by -1) on the y-axis (higher is better) against the cluster LISI on the x-axis (higher is better). The best performer on both metrics will be in the top right corner  (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')`a). In the same manner we plot the silhouette metric (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')` b). To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI).

On one extreme we have ComBat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a). The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each nearly study is has a distinct region in the UMAP (`r fig_cap(name = 'figBenchmark', display = 'cite')`b). In our scEiaD dataset we see that like ComBat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity. Scanorama, fastMNN, and bbknn prioritize cleanly separating the clusters. With our scEiaD meta-atlas insct, trVAE, and magic do not perform particularly well in batch mixing or cluster purity. Across the 11 integration methods we tested, scVI achieved the highest sumZScale for out scEiaD meta-atlas.

@@

As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c). We see that the remaining normalization techniques alter the batch correction performance, though the outcome differs across the different methods. We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance. These results demonstrate the importance of assessing performance in a rigorous manner across many parameters.

@@

```{r figBenchmark, fig.width=14, fig.height=9, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a. Example of a method (combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type). b. no batch correction cleanly separates cell types but does not mix batches (colored by study). c. sumZScale (higher is better) for each method across a variety of data normalizations. All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions. Each color is a different method. d. Boxplot of 4 different clustering resolutions for across 1000 to 10000 HVG numbers and 4 to 50 scVI latent dimensions. Open boxes are using scVI-standard and gray boxes are scVI-projection (human reference with the remaining data projected) ')

@@

To find the best set of parameters for scVI in out dataset we did a grid search across key parameters: HVG, latent dimensions, and k nearest neighbors. Furthermore we used a recent advance in scVI capability (>= version 0.8.0) adapted from scArches that allows one to build a reference model and query or project the new cells onto it [@QueryReferenceSinglecell]. We refer to this projectable model as “scVI-projection", and the previous scVI model as “scVI-standard”.  We built a scVI-projected model trained on human cells and then projected the mouse and macaque data onto it. We then compared scVI-projection against the against the previous scVI-standard across all the previously mentioned parameters. Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data (`r fig_cap(name = 'figBenchmark', display = 'cite')`d). We found the optimal overall parameters to be 5000 HVG, 8 latent dimensions, and with 5 k-nearest neighbors for the cluster finding. We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.1 and 50 nearest neighbors.

@@

As a brief case study of our ML performance, we look at the cell type labels assigned to a Shekhar et al. study where they used SMART-seq on a retinal bipolar cell enriched population of cells [@shekharComprehensiveClassificationRetinal2016] . Even though our ML algorithm was trained only on droplet-based data, our algorithm labels most of this dataset as retinal bipolar cells, with the next most common cell types being amacrine. The same result was found by Shekhkar et al. (`r supFig_cap(name = 'supFigShekharSMART', display = 'cite')`). To more generally evaluate performance we use the precision recall (PR) curve, which visualizes the ability of the model to precisely label known cells at a given confidence. The area under the PR curve (AUC) summarizes the effectiveness of the model across different cell types, with 1 being the highest performance. The xgboost model can predict rods, bipolar cells, and Müller glia with near perfect performance (`r supFig_cap(name = 'supFig_PR', display = 'cite')`). Most of the remaining cell types can be predicted with an AUC over 0.9 `r supTab_cap(name = 'supTab_AUCxgboost', display = 'cite')`. Several of the precursor cell types (photoreceptor, AC/HC, and neurogenic) were labelled with lower confidence `r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`. The next most common labels for these cell types were either other precursor cells (e.g. 100 AC/HC Precursors were labeled as Neurogenic) or the adjacent terminal cell type (e.g. 254 photoreceptor precursors were labeled as cones). Other cell types that were challenging for the model to predict were the artery, choriocapillaris, and vein. Artery, choriocapillaris, and vein are constructed from endothelial cells and we find that for all three of these, endothelial was the second most common label. Overall, our xgboost based ML model shows strong accuracy across the major cell types of the retina (overall AUC of 0.98).

```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE, warning = FALSE}
figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'a. Top genes that are differentially expressed across the major cell types of the retina (PR is short for Photoreceptor). Genes are colored by which cell type they are differentially expressed in. The dot size is proportional to the percentage of those cells that have detectable levels of the gene. The color of the dot is the log2 scaled CPM expression. b. 2D UMAP projection of scEiaD, colored by cell type (Tabula Muris data is gray). Arrows are scvelo RNA velocity. Longer arrows are cells with higher velocity (relatively more unspliced transcripts). c. Facet plot that demonstrates how each major cell type of the retina is contained within a distinct space. d. Confusion matrix of cell type prediction performance of our xgboost labeller between predicted (x axis) and known (y axis) using data withheld from the machine learner. Most of the cell types are indeed labelled as their true type. e. Faceting of 2D UMAP by species and colored by cell type demonstrate how the major cell types of the retina share space with like cell types, despite being from mouse, human, and macaque.')

@@

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 1
input[['pt_size_meta']] <- 1
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['label_toggle']] <- 2
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- NULL
ctp <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df,
                                   velocity = TRUE
)

@@

After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type. For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labeled Müller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`). Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean() %>% round()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean() %>% round()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean() %>% round()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean() %>% round()` after transferring the cell type labels.

As another check on the quality of the cell type assignments, we ran the cell and cluster independent haystack gene search and pairwise differential expression tests between the predicted cell types (see methods for further details). We show the five most differentially expressed genes for each of the major retina cell types are consistent with known retinal cell markers (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a). As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene. We expect most of the genes identified to be known in the literature. The first search is the more precise "gene AND cell type" (e.g. "PDE6H AND Cones") and the second search is the more inclusive "gene AND retina" (e.g. "PDE6H AND Retina"). Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (Supplementary File celltype_markers.tsv) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search. The 50 genes had a mean of `r top_markers %>% filter(Gene %in% dplot_markers) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search). In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 1.44 x 10^-17).

@@

## The scVI-based scEiaD UMAP projection blends batches and species while separating cell types
